\documentclass{scrartcl}
\usepackage{listings}
\usepackage{caption}
\usepackage{color}
\usepackage{booktabs}

\newcounter{nalg}[section] % defines algorithm counter for section
\renewcommand{\thenalg}{\thesection .\arabic{nalg}} %defines appearance of the algorithm counter
\DeclareCaptionLabelFormat{algocaption}{Algorithm \thenalg} % defines a new caption label as Algorithm x.y

\lstnewenvironment{algorithm}[1][] %defines the algorithm listing environment
{   
    \refstepcounter{nalg} %increments algorithm number
    \captionsetup{labelformat=algocaption,labelsep=colon} %defines the
                                                          %caption
                                                          %setup for:
                                                          %it ises
                                                          %label
                                                          %format as
                                                          %the
                                                          %declared
                                                          %caption
                                                          %label above
                                                          %and makes
                                                          %label and
                                                          %caption
                                                          %text to be
                                                          %separated
                                                          %by a ':'
    \lstset{ %this is the stype
        frame=tB,
        numbers=left, 
        numberstyle=\tiny,
        basicstyle=\scriptsize, 
        keywordstyle=\color{black}\bfseries,
        escapeinside={(*}{*)},
        keywords={,input, output, return, datatype, function, in, if, else, foreach, while, begin, end, } 
        numbers=left,
        xleftmargin=.04\textwidth,
        #1 % this is to add specific settings to an usage of this environment (for instnce, the caption and referable label)
    }
}
{}





\author{Alexander Eisl (0250266), Peter Wiedermann (0025999)}

\date{\today}


\title{Heuristic Optimization Techniques \\ Exercise 1}

\begin{document}
\maketitle


\section{Deterministic Construction Heuristic}
\label{sec:deterministic}
In the first part of this programming exercise, we develop a
deterministic construction heuristic to solve the K-page minimal
crossing problem. The following listing shows the pseudo-code of our
algorithm.

\begin{algorithm}[caption={Deterministic construction heuristic}]
    input: Graph with given spine order and set of edges
    output: Graph with new spine order and edge partition

    begin
       start with empty solution
       spine order (*$\leftarrow$*) sort(spine order)
       while solution not complete
           add edge to page that creates the fewest crossings
       end 
    end
\end{algorithm}

Our heuristic construct the solution in two parts:

\begin{itemize}
\item In the first part,
we create a new spine-order by sorting the vertices according to the
number of neighbours fo each vertex. The idea of this approach is to
get vertices with many neighbours closer together to minimize the
number of crossings these can potentially produce.

\item In the second part of the heuristic, we add edges to the pages of our 
    solution using a greedy approach. Edges are added in the order of the adjacency matrix (arbitrary).
Each edge is added to the first page where it wouldn't introduce additional crossings.
If we cannot find a page where we wouln't introduce crossings, we add the edge to
the page where we introduce the minimum number of crossings.
\end{itemize}

%%% Argue why we generate the spine-order first and not sequentially
%%% FIXME: das hinkt noch ein wenig...
We decided to generate the spine-order in the beginning and then
decide on the edge-partinioning independently. Separating the two
problems makes the algorithm simpler and computationally faster. 

% It also allows for a simple way to incorporate randomness in our construction heuristic.
We will mix spine order and edge-page allocation during our randomized
variant desribed in the section \ref{sec:randomization}.




\section{Randomization}
\label{sec:randomization}
In this section, we explain how we extend our approach to a
randomized/multi-solution heuristic. We do this by randomly
permutating a subset of the vertices on the spine. Thus, the result of
our spine-sorting algorithm is subject to random variations. 

We first start with a spine-porder as described in our deterministic
variant and apply the following randomized variant:
\begin{itemize}
\item We randomly permutate $n$ vertices $k$ times, to create $k$ different
initial spine-orders. For each of these $k$ spine-orders we can run
the same edge-allocation algorithm as described in section \ref{sec:deterministic}. 

\item We then compute the number of
crossing and -- from the list of generated solutions -- select the one
resulting in the lowest number of edge-crossings. 
\end{itemize}


%%% Why this is not random search
%Our algorithm also does not degenerate into random search, as we limit
%the number of random permutations that are allowed.

Our algorithm does not degenerate into random search because we try to
improve our deterministic heuristic approach.


\section{Solution Representation}
\input{representationTab}
Table~\ref{tab:sol} shows the basic structure of our solution representation.

\section{Results}

Table~\ref{tab:results} shows the results of our
algorithm, where we have choosen to perform $n=4$ permutations of vertices to generete different $k=5$ solutions. 

We executed the code on a desktop computer with a Core i7 Quad-Core
CPU and 24 GB of main memory. Table~\ref{tab:results} shows the
results of our randomized algorithm.
% TODO: frequenz


The solutions we generate using the random approach are
slightly better than the solutions generated by our deterministic
routine. This is not surpring, as we can make sure that the randomized
heuristic produces at least the same result as the deterministic
algorithm by including the deterministic spine-order in our set of
solution candidates.




\begin{table}[h]

\centering
\input{resultsTable}
\caption{This table shows the results of our randomized construction heuristic. The first two
columns show the mean and standard deviation of the number of crossings, the last two columns
show the mean and standard deviation of the runtime in seconds.}
\label{tab:results}
\end{table}


\end{document}
