\documentclass{scrartcl}
\usepackage{listings}
\usepackage{caption}
\usepackage{color}
\usepackage{booktabs}
\usepackage{lscape}
\usepackage{tabularx}

\newcounter{nalg}[section] % defines algorithm counter for section
\renewcommand{\thenalg}{\thesection .\arabic{nalg}} %defines appearance of the algorithm counter
\DeclareCaptionLabelFormat{algocaption}{Algorithm \thenalg} % defines a new caption label as Algorithm x.y

\lstnewenvironment{algorithm}[1][] %defines the algorithm listing environment
{   
    \refstepcounter{nalg} %increments algorithm number
    \captionsetup{labelformat=algocaption,labelsep=colon} %defines the
                                                          %caption
                                                          %setup for:
                                                          %it ises
                                                          %label
                                                          %format as
                                                          %the
                                                          %declared
                                                          %caption
                                                          %label above
                                                          %and makes
                                                          %label and
                                                          %caption
                                                          %text to be
                                                          %separated
                                                          %by a ':'
    \lstset{ %this is the stype
        frame=tB,
        numbers=left, 
        numberstyle=\tiny,
        basicstyle=\scriptsize, 
        keywordstyle=\color{black}\bfseries,
        escapeinside={(*}{*)},
        keywords={,input, output, return, datatype, function, in, if, else, foreach, while, begin, end, } 
        numbers=left,
        xleftmargin=.04\textwidth,
        #1 % this is to add specific settings to an usage of this environment (for instnce, the caption and referable label)
    }
}
{}





\author{Alexander Eisl (0250266), Peter Wiedermann (0025999)}

\date{\today}


\title{Heuristic Optimization Techniques \\ Exercise 3}

\begin{document}
\maketitle


\section{General Variable Neighborhood search}
\label{sec:general_vns}

In this programming exercise, we implement a generalized variable
neighborhood search heuristic, re-using the neighborhoods from the
previous exercise. We implement the algorithm as shown on page 11 of
the lecture slides on VNS + VDS.



%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
%%% NEIGHBORHOODS HERE
\subsection{Neighborhoods}
\label{neighborhoods}
We re-use the neighborhoods from Assignment~2 and add a parameterized node-move variant:

\begin{description}
\item[X-node-move] 
Defined as all subsets where X vertices of the initial solution are moved to other positions.

\begin{itemize}
\item Size of neigborhood: $(n-1)^2 \cdot X $
\item Objective Function: Incremental, only crossings for moved edges are recalculated.
\end{itemize}

\end{description}
 
The neighborhoods from the previous assignment:

\begin{description}
\item[1-node flip] 
    
    This neighborhood is defined as all subsets where two vertices of the initial solution are flipped.
    \begin{itemize}
        \item Size of neigborhood: $n(n-1)/2$
        \item Objective Function: Incremental, crossings from flipped vertices are subtracted and recalculated.

    \end{itemize}

	
\item[1-edge move] This neighborhood consists of all solutions where one edge is moved to a different page. 
     \begin{itemize}
        \item Size of neigborhood: $(pages-1) edges$ 
        \item Objective Function: Incremental, only crossings for moved edges are recalculated.
    \end{itemize}
   
	
\item[1-node edge move] This neighborhood consists of all solutions where the edges of
	one vertex on a specific page are moved to all different pages. 

    \begin{itemize}
        \item Size of neigborhood: $(pages-1) edges$ as worst
          case.\footnote{depending on number of edges on the respective
            node pages} %% FIXME: node pages?
        \item Objective Function: Incremental, only crossings for moved edges are re-calculated.
    \end{itemize}
    

\end{description}



\subsection{Neigborhoodstructures in General VNS}
\label{neighborhood_orders}

We define two non-overlapping sets of neigborhoods for the outer and inner loop:

\begin{description}

    \item{VNS}: used for the shaking, we decided to take the complex
      neighborhoods here, because it would be costly to search them
      completely. We use the following order:
                
                1-node-move, 2-node-move, 3-node-move
                
    \item{VND}: used for variable neighborhood descent in the inner loop. We use use the following order:
         
                1-edge-move, 1-node-edge-move

\end{description}


%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
%%% NEIGHBORHOODS HERE
\section{Results}

We executed the code on a desktop computer with a Core i7 Quad-Core
CPU with 2.67Ghz and 24 GB of main memory. 

\paragraph{}
We set a timeout of 5 minutes for the calculations.  Because we
are relying on random calculations, we provide the summary statistics
of the results over 30 independent runs.


\paragraph{}
Table~\ref{tab:resultsOverall} shows the number of crossings for the
GVNS using the order of neighborhoods as described in Section~\ref{neighborhood_orders}.
The corresponding run-times are reported in Table~\ref{tab:Timing}.

We find that for the small instances, the GVNS finds better solutions
than the local search we used in the previous exercise. The timing in
Table~\ref{tab:Timing} shows that for larger instances (except for
instance 7) the maximum run-time is usually exhausted and, as a
result, we do not find large improvements compared to the local search
in the previous exercise.

\paragraph{Order of the neighborhoods}
We run two additional experiments where we change the order of the
neighborhoods.\footnote{see Section~\ref{neighborhood_orders} for the
  original order}

Tables~\ref{tab:resultsVns} and ~\ref{tab:TimingVns} show the
results when we reverse the order of the VNS neighborhoods
used for the shaking.\footnote{Unfortunately, our script which executed
  the experiments did not include a character string specifying the
  experiment and, thus, re-used the filenames of our output files. As a result,
  we have to re-run our computations to get all the output files
  necessary for the submission on the cluster. Runtimes and crossing
  numbers were recorded separately and are correct. Because of the
  random nature of our algorithm, the crossing numbers of the solutions that we will
  submit on the cluster tomorrow might be different. Please note that
  the submission deadline on TUWEL seems to be too early (Sunday,
  00:00) and different from the one mentioned in the assignment
  (Sunday, 23:55).}

As we can see, the reverse order of the VNS neighborhoods improves the minimum number of crossings. If we are looking
at the timing results of instance 7 in Table~\ref{tab:TimingVns}, we see that the variance in the runtime is significantly
higher than in the original order.
%% TODO: removed; I wouldn't call hitting the max run-time 'stability'
%% This means that we get a better minimum for the cost of stability of the solution,
%%originating in the fact that we start exchaning 3 random vertices, than just 1 as in the original order.


\begin{table}[!H]
  \centering
  \scriptsize
  \input{crossingsTab}
\caption{Number of crossings using default neighborhood order}
\label{tab:resultsOverall}
\end{table}

\begin{table}[!H]
  \centering
  \scriptsize
  \input{crossingsTab-stoch-order}
\caption{Number of crossings using reverse order of VNS neighborhood}
\label{tab:resultsVns}
\end{table}

\begin{table}[!H]
  \centering
  \scriptsize
  \input{crossingsTab-det}
\caption{Number of crossings using reverse order of VND neighborhood}
\label{tab:resultsVnd}
\end{table}

\clearpage

\begin{table}[H]
  \centering
  \scriptsize
  \input{timeTab}
\caption{Runtime in seconds using default neighborhood order}
\label{tab:Timing}
\end{table}

\begin{table}[H]
  \centering
  \scriptsize
  \input{timeTab-stoch-order}
\caption{Runtime in seconds using reverse order of VNS neighborhood}
\label{tab:TimingVns}
\end{table}

\begin{table}[H]
  \centering
  \scriptsize
  \input{timeTab-det-order}
\caption{Runtime in seconds using reverse order of VND neighborhood}
\label{tab:TimingVnd}
\end{table}







\end{document}
