\documentclass{scrartcl}
\usepackage{listings}
\usepackage{caption}
\usepackage{color}
\usepackage{booktabs}
\usepackage{lscape}
\usepackage{tabularx}

\newcounter{nalg}[section] % defines algorithm counter for section
\renewcommand{\thenalg}{\thesection .\arabic{nalg}} %defines appearance of the algorithm counter
\DeclareCaptionLabelFormat{algocaption}{Algorithm \thenalg} % defines a new caption label as Algorithm x.y

\lstnewenvironment{algorithm}[1][] %defines the algorithm listing environment
{   
    \refstepcounter{nalg} %increments algorithm number
    \captionsetup{labelformat=algocaption,labelsep=colon} %defines the
                                                          %caption
                                                          %setup for:
                                                          %it ises
                                                          %label
                                                          %format as
                                                          %the
                                                          %declared
                                                          %caption
                                                          %label above
                                                          %and makes
                                                          %label and
                                                          %caption
                                                          %text to be
                                                          %separated
                                                          %by a ':'
    \lstset{ %this is the stype
        frame=tB,
        numbers=left, 
        numberstyle=\tiny,
        basicstyle=\scriptsize, 
        keywordstyle=\color{black}\bfseries,
        escapeinside={(*}{*)},
        keywords={,input, output, return, datatype, function, in, if, else, foreach, while, begin, end, } 
        numbers=left,
        xleftmargin=.04\textwidth,
        #1 % this is to add specific settings to an usage of this environment (for instnce, the caption and referable label)
    }
}
{}





\author{Alexander Eisl (0250266), Peter Wiedermann (0025999)}

\date{\today}


\title{Heuristic Optimization Techniques \\ Exercise 2}

\begin{document}
\maketitle


\section{Local Search}
\label{sec:deterministic}
In this programming exercise, we improve the results of the
construction heuristic developed in the previous exercise by using a
local search. In the following subsection, we will first explain the
neighborhoods that we define and then outline the step functions used.

%%% brauchen wir sowas?
\begin{algorithm}[caption={Local search}]
    input: Graph 
    output: Improved solution
    begin
    	sol (*$\leftarrow$*) calculate initial solution
	repeat:
		choose sol' (*$\in$*) N(x) 
		if (*$f(sol')$*) (*$\leq$*) f(sol) then
			sol (*$\leftarrow$*) sol'
	until stopping criteria satisfied
    end
\end{algorithm}


%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
%%% NEIGHBORHOODS HERE
\subsection{Neighborhoods}
\label{neighborhoods}
We use the following neighbourhoods:

\begin{description}
\item[1-node flip] 
    
    This neighbourhood is defined as all subsets where two vertices of the initial solution are flipped. 
    \begin{itemize}
        \item Size of neigborhood: $n(n-1)/2$
        \item Objective Function: Incremental, crossings from flipped vertices are subtracted and recalculated.

    \end{itemize}

	
\item[1-edge move] This neighbourhood consists of all solutions where one edge is moved to a different page. 
     \begin{itemize}
        \item Size of neigborhood: $(pages-1) edges$ 
        \item Objective Function: Incremental, only crossings for moved edges are recalculated.
    \end{itemize}
   
	
\item[1-node edge move] This neighbourhood consists of all solutions where the edges of
	one vertex on a specific page are moved to all different pages. 

    \begin{itemize}
        \item Size of neigborhood: $(pages-1) edges$ as worst case \footnote{depending on number of edges on respective node pages}.
        \item Objective Function: Incremental, only crossings for moved edges are recalculated.
    \end{itemize}
    

\end{description}

%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
%%% STEPFUNCTIONS HERE
\subsection{Stepfunctions}

We implemented the following thre stepfunctions:

\begin{description}
\item[first] This step-function selects the first better solution found in the neighborhood.
\item[best] This step-function iterates over all elements of the neighbourhood and chooses the best one.
\item[random] This step-function selects a random element of the neighbourhood. 
\end{description}


%%% REMOVED, I don't think this is necessary
%%\section{Solution Representation}
%%The same as in exercise 1. Do we need this subsection?


%%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5
%%% NEIGHBORHOODS HERE
\section{Results}

We executed the code on a desktop computer with a Core i7 Quad-Core
CPU with 2.67Ghz and 24 GB of main memory. Table~\ref{tab:results}
shows the results of our local search. 

We consider a local optimum if we cannot find a better solution after 10 iterations.
We have set a timout of 5 minutes for the calculations.

\paragraph{Best neighborhood step function strategy}

We compare the results for
three different neighboorhoods and all the implemented step-functions.

The best performing neighboorhood structure is \texttt{1-node-flip}, as
shown in table~\ref{tab:results}.
For smaller instances (1-5), the \texttt{1-node-flip} in combination with 
the \texttt{first} delivers the best results.
For bigger instances (6-10) the \texttt{1-node-flip}  finds 
better solutions in combination with the \texttt{random} stepfunction.



A very interesting result is the performance of the \texttt{best} stepfunction
for the \texttt{1-node-flip}. For the given instances the \texttt{best} function
reaches a local optimum very quickly while \texttt{random} and \texttt{first} 
are calculating more iterations or even calculate until timeout for bigger
instances and can find better solutions beside the local optimum of the \texttt{best} 
variant.
%
%Overall, the combination of the 1-node-flip neighborhood and the best
%step-function shows the best improvements. For example, it finds a
%solution with only $12$ edge-crossings for the instance automatic-4.
%Not surprisingly, it also has the highest run-time.




\paragraph{Initial Solution}
The choice of the initial solution is crucial to achieve results in reasonable calculation time.
Using random initialization can easily degenerate into random search or end in a 
local optimum with less improvement to a random solution itself.



% TODO:
%When we compare the results with a random initial solution we could
%measure an avg increase of XXX crossings 
%\footnote{measured with 1-node flip and first stepfunction}.


% OLD:

%It is important to use a good initial solution
%. First, because this
%helps to end up in a good local solution.\footnote{For example, the
%  results shown on the AC cluster indicate that it should be possible
%  to find a solution with zero crossings. However, our local solution
%  is not able to find them given the initial solutions provided by our
%  construction heuristic.} Second, it can reduce the run-time of the
%algorithm because it might reduce the number of steps taken to reach
%the local optimum.


\paragraph{Neighborhoods}
Incremental objective functions implemented as described in section \ref{neighborhoods}
are performing equally well for all 3 neighborhoods.

Subsequent (possibly random) moves in our neighbourhoods can only reach
valid solutions in the search space because every solution in our neighborhoods
is valid.

\paragraph{Reaching local optimum}
If a local optimum is reached it is found in less than 10 iterations in most cases.
As you can see in table ~\ref{tab:results} there are only a few exceptions.
The results with iterations=$-1$ didn't reach a local optimum and ran into our
timeout condition.





%% Q2: Can subsequent (possibly random) moves in your neighbourhood(s) reach
%% any valid solution in the search space?
%% A2: 
Some notes about our neighborhoods:

\begin{itemize}
  \item For many graphs, we do not find an improvement. So with respect to our neighborhood we
  seem to be in a local optimum. 
  \item It seems to be important to use the ``best''
    step-function. Iteratively using the ``first'' does not yield the
    same optimum. Once we choose an element of the neighborhood, other
    neighbors of the initial neighborhood cannot be reached
    anymore.\footnote{This seems intuitiv given the definitions of our
      neighborhood}
\end{itemize}





\begin{landscape}
\begin{table}
  \scriptsize
  \input{resultsTab}
\caption{Local search results. For each instance, we show the number of crossings (first row), the iteration 
needed to reach the local optimum (second row, first value) and the run-time of our algorithm (second row, second value).
If no local optimum was found the number of iterations is -1. }
\label{tab:results}
\end{table}
\end{landscape}



\end{document}
